---
title: "Falkor Control Script"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(tidyverse)
library(data.table)
library(pbapply)
library(xcms)
register(BPPARAM = SnowParam(tasks = length(ms_files), progressbar = TRUE))
source("scripts/functions.R")
```

## Setup and metadata things

The raw data is stored on the Synology drive in the Ingalls Lab as .raw file types. This data isn't super useful in its proprietary format, so it's easiest to convert those .raw files into .mzMLs with the "RunMsconvert.cmd" script in the project working directory. We write these files into the "mzMLs" folder in the working directory since they're also accessed by MSDIAL.

File contents (as of June 24, 2020):

```
msconvert Z:\1_QEdata\LTC\DATA\HILIC\190718_DepthProfiles_FK180310\*.raw --mzML --filter "peakPicking true 1-" --filter "polarity positive" -z -o mzMLs/pos/
msconvert Z:\1_QEdata\LTC\DATA\HILIC\HILIC_2017\170706_Gradients2.0_CruiseFilters_2\170706_Blk_Blk0p2_?.raw --mzML --filter "peakPicking true 1-" --filter "polarity positive" -z -o mzMLs/pos/
msconvert Z:\1_QEdata\LTC\DATA\HILIC\190718_DepthProfiles_FK180310\MSMS\*DDApos*.raw --mzML --filter "peakPicking true 1-" --filter "polarity positive" -z -o mzMLs/pos/MSMS/
msconvert Z:\1_QEdata\LTC\DATA\HILIC\190718_DepthProfiles_FK180310\*.raw --mzML --filter "peakPicking true 1-" --filter "polarity negative" -z -o mzMLs/neg/
msconvert Z:\1_QEdata\LTC\DATA\HILIC\HILIC_2017\170706_Gradients2.0_CruiseFilters_2\170706_Blk_Blk0p2_?.raw --mzML --filter "peakPicking true 1-" --filter "polarity negative" -z -o mzMLs/neg/
msconvert Z:\1_QEdata\LTC\DATA\HILIC\190718_DepthProfiles_FK180310\MSMS\*DDAneg*.raw --mzML --filter "peakPicking true 1-" --filter "polarity negative" -z -o mzMLs/neg/MSMS/
```

The code above calls msconvert, a command-line utility from Proteowizard, to convert the files into mzML format. It also splits them into positive and negative mode and separates out the MSMS information into a sub-folder. Per Katherine's recommendation, I'm using blanks from her Gradients 2.0 project rather than the ones Laura added for me.

We also need the metadata that goes with these files, containing information about treatment type, which are blanks, injection volume, etc. It's also good to pull down the standards file from the lab Github during this setup phase. The small script below does this by accessing the file names, the instrument-produced sample key, and the internet.

```{r Raw datagrab}
polarity <- "pos"

# Metadata
ms_files <- "../mzMLs/" %>%
  paste0(polarity) %>% #Same in both polarities, but gotta choose one!
  list.files(pattern = ".mzML", full.names = TRUE) %>%
  normalizePath() %>%
  `[`(!grepl("Fullneg|Fullpos|QC-KM1906", x = .))
metadframe <- data.frame(
  fileid=basename(ms_files),
  depth=regmatches(regexpr(pattern = "Std|Poo|Blk|DCM|25m", ms_files), x = ms_files),
  station=regmatches(regexpr(pattern = "Std|Poo|Blk|S62|S64|S77|S80", ms_files), 
                     x = ms_files)
)
station_spindirs <- c(Blk="Blk", S62="Cyclone", S64="Cyclone", 
                      S77="Anticyclone", S80="Anticyclone",
                      Poo="Poo", Std="Std")
metadframe$spindir=station_spindirs[metadframe$station]
station_times <- c(Blk="Blk", S62="Morning", S64="Afternoon", 
                   S77="Morning", S80="Afternoon",
                   Poo="Poo", Std="Std")
metadframe$time=station_times[metadframe$station]
write.csv(x = metadframe, row.names = FALSE, file = "falkor_metadata.csv")
knitr::kable(head(metadframe))



# Sample key information
sample_key <- read.csv(paste0("Z:\\1_QEdata\\LTC\\DATA\\HILIC\\190718_Depth",
                              "Profiles_FK180310\\Sample.Key.HILIC.csv"),
                       skip = 1)
falkor_key <- sample_key %>%
  filter(grepl(pattern = "FK|Std", sample_key$Sample.Name)) %>%
  select(Sample.Name, Inj.vol) %>%
  add_row(Sample.Name="170706_Blk_Blk0p2_1", Inj.vol=1) %>%
  add_row(Sample.Name="170706_Blk_Blk0p2_2", Inj.vol=1) %>%
  add_row(Sample.Name="190715_Blk_KM1906U14-Blk_C", Inj.vol=1) %>%
  mutate(fileid=paste0(Sample.Name, ".mzML")) %>%
  select(-Sample.Name, inj_vol=Inj.vol) %>%
  arrange(fileid)
write.csv(falkor_key, file = "falkor_sample_key.csv", row.names = FALSE)
knitr::kable(head(falkor_key))


# Standards information
raw_stans <- read.csv(paste0("https://raw.githubusercontent.com/",
                             "IngallsLabUW/Ingalls_Standards/master/",
                             "Ingalls_Lab_Standards_NEW.csv"))
falkor_stans <- raw_stans %>%
  filter(Column=="HILIC") %>%
  mutate(polarity=tolower(gsub(pattern = "HILIC", "", .$Fraction1))) %>%
  select(compound_type=Compound.Type, compound_name=Compound.Name, 
         formula=Emperical.Formula, rt=RT..min., mz=m.z, ionization_form,
         charge=z, kegg_id=C0, polarity)
write.csv(falkor_stans, file = "falkor_stans.csv", row.names = FALSE)
knitr::kable(head(falkor_stans))


# Create folders in which to store intermediate and final data
pretty_folder <- paste0(polarity, "_pretty/")
if(!dir.exists(pretty_folder)){dir.create(pretty_folder)}
intermediate_folder <- paste0(polarity, "_intermediate/")
if(!dir.exists(intermediate_folder)){dir.create(intermediate_folder)}
```

## Peakpicking

**Obviously** the most important step, you can't do anything else without first finding things to look at and talk about. The script below, when sourced, runs XCMS's centWave peakpicking. It also calculates an improved signal-to-noise and metric of Gaussian-ness that seems to sift through the noise more accurately than the default signal-to-noise ratio, which has known bugs (https://doi.org/10.1021/acs.analchem.7b01069). Finally, it performs retention time correction and peak correspondence (grouping peaks across files).

It takes in the metadataframe created above as well as the list of paths to the .mzML files and returns a (rather large) data frame full of peaks. 

```{r Peakpicking}
falkor_metadata <- read.csv("falkor_metadata.csv")
ms_files <- falkor_metadata$fileid
qscore_threshold <- 25

source("scripts/peakpicking.R")

saveRDS(xdata_filled, file = paste0(intermediate_folder, "current_xdata_filled.rds"))
print(xdata_filled)
write.csv(raw_peaks, file = paste0(intermediate_folder, "raw_peaks.csv"), row.names = FALSE)
knitr::kable(head(raw_peaks))
```

## De-isotoping and de-adducting

```{r deisoadduct}
xdata_filled <- readRDS(file = paste0(intermediate_folder, "current_xdata_filled.rds"))
raw_peaks <- read.csv(file = paste0(intermediate_folder, "raw_peaks.csv"))

source("scripts/deisoadduct.R")
```

## Finding B-MISs

```{B-MIS}
xdata_filled <- readRDS(paste0(intermediate_folder, "current_xdata_filled.rds"))
raw_peaks <- read.csv(paste0(intermediate_folder, "raw_peaks.csv"))
addiso_features <- read.csv(paste0(pretty_folder, "addiso_features.csv"))
is_peak_iso <- read.csv(paste0(intermediate_folder, "is_peak_iso.csv"))
complete_peaks <- read.csv(paste0(intermediate_folder, "complete_peaks.csv"))
bionorm_values <- "XCMS/Sample.Key.Falkor.Manual.csv" %>%
  read.csv() %>%
  select(file_name="ï..Sample.Name", norm_vol="Bio.Normalization")
cut.off <- 0.4 #Necessary improvement for "acceptable"
cut.off2 <- 0.1 #If RSD already below, skip B-MIS

source("scripts/bmisscript.R")
```

